Inattentive responding can induce spurious associations between task behavior and symptom measures

### 2023-04-03 manuscripts/main.tex
checkpointing manuscript progress

To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder (see Appendix C for details). Specifically, N=45 psychiatric patients (confirmed to meet criteria for a diagnosis of major depressive disorder through a structured clinical interview) and N=20 healthy controls, all recruited through the Rutgers-Princeton Center for Computational Cognitive Neuropsychiatry (i.e., personal recruitment, not on online labor platforms), completed a series of self-report symptom measures, online, on their computers from the comfort of their homes. In total, 16 of 65 (24.6\%) participants failed one or more attention checks. Subdivided by group, 6 of 20 (30\%) healthy participants and 10 of 45 (22\%) psychiatric participants were flagged for C/IE responding.

@@

In this manuscript, we highlighted a particular set of circumstances, common in computational psychiatry research done on large online samples, in which spurious correlations may arise between task behavior and self-reported symptomology. When the ground-truth prevalence of a symptom is low in the general population, participants who respond carelessly on measures assessing this symptom may erroneously appear as symptomatic. A less careful pattern of responding on tasks used to measure cognitive constructs can then masquerade as a correlation between alteration in these constructs and symptom dimensions. We found repeated evidence for this pernicious pattern in participants recruited from two popular online labor platforms. False-positive rates for these spurious correlations \emph{increased} with sample size, because the correlations are due to measurement bias, not measurement noise. Importantly, we found that screening on task behavior alone was insufficient to identify participants engaging in C/IE responding and avoid the false-positive correlations. Unfortunately, a literature review identified this type of screening as the most common practice in online computational psychiatry studies. We recommend instead to screen and exclude participants based on responding on surveys, a practice that abolished many spurious behavior-symptom correlations in our data.

@@

Viewed from the perspective of effort expenditure, our results suggest that participants appraised the cost/benefit trade-off differently for behavioral tasks and self-report surveys. Specifically, we found that only 7\% of participants were at chance-level performance in our task, compared to more than 22\% of participants who failed one or more attention-check items in the self-report surveys (a finding that qualitatively replicated in a second study involving a different task). Moreover, different measures of C/IE responding were weakly or not at all correlated between task behavior and self-report responses. This suggests the motivation for effortful responding was greater in the behavioral task, though precisely why is unclear. One possibility is that we gave participants a monetary incentive for attentive responding only during the task (a common practice, according to our literature review). A second possibility is that participants expected fewer consequences for C/IE responding during the self-report surveys, a reasonable assumption in light of how infrequently previous experiments have screened self-report data. Alternatively, participants may have found the gamified behavioral task more engaging or the self-report inventory more tedious. Regardless of the reason, this discrepancy reinforces our observations concerning the inadequacy of behavioral-task screening as a stand-alone method for identifying C/IE responding. Since, in general, participants may appraise costs and benefits of effortful responding differently for behavioral tasks and self-report surveys, screening for C/IE responding on one data modality may in general be unsuitable for identifying it in the other. We therefore recommend screening on each component of an experiment.

@@

One objection to the rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias. That is, to the extent that C/IE responding might reflect symptoms common to psychopathology (e.g., low motivation, effort avoidance, inattentiveness), rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants. To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder. Though our final sample was small, we did not find evidence for the proposition that patients were more likely to fail attention checks than healthy controls (if anything, non-psychiatric participants were more likely to be flagged by C/IE screening). These results provide preliminary evidence that rigorous C/IE screening is unlikely to result in overcontrol bias.

@@

There are several notable limitations to this proof-of-concept study. We used a small set of screening measures, and did not employ other recommended procedures (e.g., logging each key/mouse interaction during survey administration to detect form-filling software or other forms of speeded responding \cite{buchanan2018methods}). Thus, we cannot be confident that all of the flagged participants were indeed engaging in C/IE responding; similarly, we cannot be certain that we correctly excluded all participants engaged in C/IE responding. We studied behavior-symptom correlations for only two tasks and two short self-report symptom batteries. It remains to be seen how generalizable our findings are, although our study design was inspired by experiments prevalent in the online computational psychiatry literature. As suggested above, future studies may find greater correspondence between task and self-report screening measures for more difficult behavioral experiments. Finally, we should note that, unlike previous studies in which some participants were explicitly instructed to respond carelessly \cite{buchanan2018methods}, we do not have access to ``ground truth'' regarding which participants were engaging in C/IE responding. Future work testing the efficacy of different screening metrics for identifying instructed C/IE responding may help to identify some of the issues that we have identified here.

The present study highlights the need for future work on the prevalence of C/IE responding in online samples and its interactions with task-symptom correlations. Many open questions remain, including under what conditions task- and symptom-screening measures might better correspond, what screening measures are most effective and when, and under what conditions spurious correlations are more likely to arise. One especially pressing question is how sample size affects the likelihood of obtaining spurious correlations. The results of a bootstrapping analysis in our data suggest that false positive rates are likely to increase with sample size. As computational psychiatry studies move towards larger samples to characterize heterogeneity in symptoms (and to increase statistical power), it will be important to understand how sample size may exaggerate the effects of systematic error.

\section*{Conclusions}

@@

\begin{figure}[h]
    % \includegraphics[width=16cm]{../figures/main_01.png}
    \centering
    \captionsetup{width=0.88\textwidth}
    \caption{Simulated example of how spurious behavior-symptom correlations can arise when symptom endorsement is rare. \textit{Left:} When symptoms are moderately common in the general population, C/IE respondents (blue) are indistinguishable from attentive participants (red) in self-report measures (x-axis, marginal distribution shown on top). Despite the lower task performance of C/IE respondents (y-axis), no correlation arises between symptom scores and task performance (dots are participants drawn from the shown distributions, with 15\% C/IE participants; dashed line shows the (lack of) correlation.) \textit{Right:} When symptoms are rare in the general population, careless respondents appear symptomatic in self-report measures. As a result, self-report symptom scores show a significant correlation with task performance.}
    \label{fig:simulation}
\end{figure}

\clearpage
\begin{figure}[h]
    % \includegraphics[width=16cm]{../figures/main_02a.png}
    \centering
    \captionsetup{width=0.88\textwidth}
    \caption{Raincloud plots of total symptom scores in attentive (red) and C/IE (blue) participants. Each colored dot represents the symptom score for one participant. Black circles: average score within each group (error bars denote 95\% bootstrap confidence interval). Shaded plots: estimated distribution of responses for each group of participants. The scales are ordered approximately according to their estimated skew (see Table 2) from top-left (7-up) to bottom-right (PSWQ). The average level of symptom endorsement is most markedly different between groups in measures with lowest overall rates of endorsement.}
    \label{fig:distributions}
\end{figure}

\clearpage
\begin{figure}[h]
    % \includegraphics[width=16cm, trim=0 0 3cm 0]{../figures/main_02b.png}
    \centering
    \captionsetup{width=0.88\textwidth}
    \caption{Similarity of task and self-report data screening measures. Each tile corresponds to the Spearman rank correlation (left) and Dice similarity coefficient (right) between two screening measures. Acronyms: INF = infrequency item; ISD = inter-item standard deviation; REL = personal reliability; MAH = Mahalanobis distance; READ = reading time; VAR = choice variability; ACC = choice accuracy; WSLS = win-stay lose-shift rate; RT = suspicious response times. Similarity scores have been thresholded after correcting for multiple comparisons. Numbers denote the strength of statistically significant correlations. Cross-modality correlations between task-behavior (left) and infrequency-item self-report measures (bottom) are in the dashed rectangle.}
    \label{fig:correspondence}
\end{figure}

\clearpage
\begin{figure}[h]
    % \includegraphics[width=17cm]{../figures/main_03a.png}
    \centering
    \captionsetup{width=0.88\textwidth}
    \caption{Absolute Spearman rank correlations between task behavior (y-axis) and symptom measures (x-axis) under different regimes of data screening and participant exclusions. Only statistically significant correlations are shown ($p<0.05$ not corrected for multiple comparisons; all signed correlations are shown in Supplementary Figure \ref{fig:correlations_s1}). Black Xs indicate significant correlations ablated under screening. No Screening = no exclusions; Accuracy Only = exclusions based on chance-level performance in the reversal-learning task; Infrequency Only = exclusions based on invalid or improbable responses to infrequency items; Both Types = exclusions based on the previous two measures. Acronyms: Acc = choice accuracy; Pts = total points earned; WS = win-stay rate; LS = lose-shift rate; Pers = perseveration errors; $\beta$ = inverse temperature; $\eta_p$ = positive learning rate; $\eta_n$ = negative learning rate; $\kappa$ = learning rate asymmetry.}
    \label{fig:correlations}
\end{figure}

\clearpage
\begin{figure}[h]
    % \includegraphics[width=16cm]{../figures/main_03b.png}
    \centering
    \captionsetup{width=0.88\textwidth}
    \caption{False positive rates for spurious correlations \emph{increase} with sample size. \textit{Left:} Bootstrapped Spearman correlations between learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. The grey error bars indicate 95\% bootstrap confidence intervals. The thick dashed lines indicate the threshold for statistical significance for the Spearman correlation at the corresponding sample size. Markers are jittered along the x-axis for legibility. \textit{Right:} False positive rates for learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. False positive rate was calculated as the proportion of bootstrap samples in which the correlation between $\kappa$ and 7-down was statistically significant. The horizontal dotted line denotes the expected false positive rate at $\alpha = .05$.}
    \label{fig:bootstrap}
\end{figure}

>>>>

To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder (see Appendix C for details). Specifically, N=45 psychiatric patients (confirmed to meet criteria for a diagnosis of major depressive disorder through a structured clinical interview) and N=20 healthy controls, all recruited through the Rutgers-Princeton Center for Computational Cognitive Neuropsychiatry (i.e., not on online labor platforms), completed a series of self-report symptom measures, online, on their computers from the comfort of their homes. In total, 16 of 65 (24.6\%) participants failed one or more attention checks. Subdivided by group, 6 of 20 (30\%) healthy participants and 10 of 45 (22\%) psychiatric participants were flagged for C/IE responding.

@@

In this manuscript, we highlighted a particular set of circumstances, common in computational psychiatry research done on large online samples, in which spurious correlations may arise between task behavior and self-reported symptomology. When the ground-truth prevalence of a symptom is low in the general population, participants who respond carelessly on measures assessing this symptom may erroneously appear as symptomatic. A less careful pattern of responding on tasks used to measure cognitive constructs can then masquerade as a correlation between alteration in these constructs and symptom dimensions. We found repeated evidence for this pernicious pattern in participants recruited from two popular online labor platforms. False-positive rates for these spurious correlations \emph{increased} with sample size, because the correlations are due to measurement bias, not measurement noise. Importantly, we found that screening on task behavior alone was insufficient to identify participants engaging in C/IE responding and prevent some false-positive correlations. Unfortunately, a literature review identified this type of screening as the most common practice in online computational psychiatry studies. We recommend instead to screen and exclude participants based on responding on surveys, a practice that abolished many spurious behavior-symptom correlations in our data.

@@

Viewed from the perspective of effort expenditure, our results suggest that participants appraised the cost/benefit trade-off differently for behavioral tasks and self-report surveys. Specifically, we found that only 7\% of participants in the first study were at chance-level performance in the task, compared to more than 22\% of participants who failed one or more attention-check items in the self-report surveys (a finding that qualitatively replicated in a second study involving a different task). Moreover, different measures of C/IE responding were weakly or not at all correlated between task behavior and self-report responses. This suggests the motivation for effortful responding was greater in the behavioral task, though precisely why is unclear. One possibility is that we gave participants a monetary incentive for attentive responding only during the task (a common practice, according to our literature review). A second possibility is that participants expected fewer consequences for C/IE responding during the self-report surveys, a reasonable assumption in light of how infrequently previous experiments have screened self-report data. Alternatively, participants may have found the gamified behavioral task more engaging or the self-report inventory more tedious. Regardless of the reason, this discrepancy reinforces our observations concerning the inadequacy of behavioral-task screening as a stand-alone method for identifying C/IE responding. Since, in general, participants may appraise costs and benefits of effortful responding differently for behavioral tasks and self-report surveys, screening for C/IE responding on one data modality may in general be unsuitable for identifying it in the other. We therefore recommend screening on each component of an experiment.

@@

One objection to the rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias. That is, to the extent that C/IE responding might reflect symptoms common to psychopathology (e.g., low motivation, effort avoidance, inattentiveness), rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants. To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder. Though our final sample was small, we did not find evidence that depressed patients were more likely to fail attention checks than healthy controls (if anything, non-psychiatric participants were more likely to be flagged by C/IE screening). These results provide preliminary evidence that rigorous C/IE screening is unlikely to result in overcontrol bias. However, further research with larger samples is necessary to validate attention checks in depressed and other patient populations.

@@

There are several notable limitations to this proof-of-concept study. We used a small set of screening measures, and did not employ other recommended procedures (e.g., logging each key/mouse interaction during survey administration to detect form-filling software or other forms of speeded responding \cite{buchanan2018methods}). Thus, we cannot be confident that all of the flagged participants were indeed engaging in C/IE responding; similarly, we cannot be certain that we correctly excluded all participants engaged in C/IE responding. We studied behavior-symptom correlations for only two tasks and two sets of self-report instruments. It remains to be seen how generalizable our findings are, although our study design was inspired by experiments prevalent in the online computational psychiatry literature. As suggested above, future studies may find greater correspondence between task and self-report screening measures for more difficult behavioral experiments. Finally, we should note that, unlike previous studies in which some participants were explicitly instructed to respond carelessly \cite{buchanan2018methods}, we do not have access to ``ground truth'' regarding which participants were engaging in C/IE responding. Future work testing the efficacy of different screening metrics for identifying instructed C/IE responding may help to identify some of the issues that we have identified here.

The present study highlights the need for future work on the prevalence of C/IE responding in online samples and its interactions with task-symptom correlations. Many open questions remain, including under what conditions task- and symptom-screening measures might better correspond, what screening measures are most effective and when, and under what conditions spurious correlations are more likely to arise. For example, in the present study, we found that screening on task behavior alone was insufficient to prevent putatively spurious correlations for one task (reversal learning) but not for another (two-step task). This discrepancy may reflect differences in the tasks (e.g., the two-step task may be more challenging and thus more sensitive to C/IE responding) or differences in the screening measures (e.g., choice accuracy across 90 trials may be a noisier measure than win-stay lose-shift choice behavior across 200 trials). One especially pressing question is how sample size affects the likelihood of obtaining spurious correlations. The results of a bootstrapping analysis in our data suggest that false positive rates are likely to increase with sample size. As computational psychiatry studies move towards larger samples to characterize heterogeneity in symptoms (and to increase statistical power), it will be important to understand how sample size may exaggerate the effects of systematic error. It will also be important to understand how this is moderated by overall C/IE responding rates, which we observed to vary across platforms and time, and will presumably continue to change with evolving labor platform and researcher screening practices.

% \section*{Conclusions}

@@

\begin{figure}[h]
    \includegraphics[width=16cm]{../figures/main_01.png}
    \centering
    \captionsetup{width=0.88\textwidth}
    \caption{Simulated example of how spurious behavior-symptom correlations can arise when symptom endorsement is rare. \textit{Left:} When symptoms are moderately common in the general population, C/IE respondents ($N=15$; blue) are indistinguishable from attentive participants ($N=85$; red) in self-report measures (x-axis, marginal distribution shown on top). Despite the worse task performance of C/IE respondents (y-axis), no correlation arises between symptom scores and task performance (dots are participants drawn from the shown distributions, with 15\% C/IE participants; dashed line shows the (lack of) Spearman rank correlation.) \textit{Right:} When symptoms are rare in the general population, careless respondents appear symptomatic in self-report measures. As a result, self-report symptom scores show a significant two-sided Spearman rank correlation with task performance. }
    \label{fig:simulation}
\end{figure}

\clearpage
\begin{figure}[h]
    \includegraphics[width=16cm]{../figures/main_02a.png}
    \centering
    \captionsetup{width=0.88\textwidth}
    \caption{Raincloud plots of total symptom scores in attentive ($N=301$; red) and C/IE ($N=85$; blue) participants. Each colored dot represents the symptom score for one participant. Black circles: average score within each group (error bars denote 95\% bootstrap confidence interval). Shaded plots: estimated distribution of responses for each group of participants. The scales are ordered approximately according to their estimated skew (see Table 2) from top-left (7-up) to bottom-right (PSWQ). The average level of symptom endorsement is most markedly different between groups in measures with lowest overall rates of endorsement.}
    \label{fig:distributions}
\end{figure}

\clearpage
\begin{figure}[h]
    \includegraphics[width=16cm, trim=0 0 3cm 0]{../figures/main_02b.png}
    \centering
    \captionsetup{width=0.88\textwidth}
    \caption{Similarity of task and self-report data screening measures. Each tile corresponds to the Spearman rank correlation (left) and Dice similarity coefficient (right) between two screening measures across $N=386$ participants. Acronyms: INF = infrequency item; ISD = inter-item standard deviation; REL = personal reliability; MAH = Mahalanobis distance; READ = reading time; VAR = choice variability; ACC = choice accuracy; WSLS = win-stay lose-shift rate; RT = suspicious response times. Similarity scores have been thresholded after correcting for multiple comparisons. Numbers denote the strength of statistically significant correlations. Cross-modality correlations between task-behavior (left) and infrequency-item self-report measures (bottom) are in the dashed rectangle.}
    \label{fig:correspondence}
\end{figure}

\clearpage
\begin{figure}[h]
    \includegraphics[width=17cm]{../figures/main_03a.png}
    \centering
    \captionsetup{width=0.88\textwidth}
    \caption{Absolute Spearman rank correlations between task behavior (y-axis) and symptom measures (x-axis) under different regimes of data screening and participant exclusions. (A) No Screening = no exclusions ($N=386$). (B) Accuracy Only = exclusions based on chance-level performance in the reversal-learning task ($N=352$). (C) Infrequency Only = exclusions based on invalid or improbable responses to infrequency items ($N=301$). (D) Both Types = exclusions based on the previous two measures ($N=283$). Only statistically significant correlations are shown ($p<0.05$ not corrected for multiple comparisons; all signed correlations are shown in Supplementary Figure \ref{fig:correlations_s1}). Black Xs indicate significant correlations ablated under screening. Acronyms: Acc = choice accuracy; Pts = total points earned; WS = win-stay rate; LS = lose-shift rate; Pers = perseveration errors; $\beta$ = inverse temperature; $\eta_p$ = positive learning rate; $\eta_n$ = negative learning rate; $\kappa$ = learning rate asymmetry.}
    \label{fig:correlations}
\end{figure}

\clearpage
\begin{figure}[h]
    \includegraphics[width=16cm]{../figures/main_03b.png}
    \centering
    \captionsetup{width=0.88\textwidth}
    \caption{False positive rates for spurious correlations \emph{increase} with sample size. \textit{Left:} Spearman correlations and 95\% bootstrap confidence intervals between learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. The thick dashed lines indicate the threshold for statistical significance for the Spearman correlation at the corresponding sample size. Markers are jittered along the x-axis for legibility. \textit{Right:} False positive rates for learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. False positive rate was calculated as the proportion of bootstrap samples in which the correlation between $\kappa$ and 7-down was statistically significant. The horizontal dotted line denotes the expected false positive rate at $\alpha = .05$.}
    \label{fig:bootstrap}
\end{figure}
